---
title: "Racial Segregation and Black-White Longevity Disparities^[I would like to thank Casey Breen, Katharina Geppert, Matt Hall, Nathan Ly, Álvaro Padilla-Pozo, Peter Rich, Cristobal Young, Erin York Cornwell, David F. Warner, participants of the Cornell Inequality Discussion Group, SpacePop Working Group, The University of Washington Computational Demography Working Group, and PAA 2025 participants for helpful feedback on earlier versions of this paper. This work was supported by funding from the Cornell Sage Fellowship. A replication package for this paper is available at this link: https://github.com/theloniousgoerz/bw_segregation_longevity.]"
author: |
  | Thelonious Goerz$^{12}$
  | $^{1}$Department of Sociology, Cornell University
  | $^{2}$Brooks School of Public Policy, Cornell University
  | tlg95@cornell.edu
date: "Draft date: `r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document:
    number_sections: yes
linestretch: 1.5
fig_width: 5
fig_height: 2.5
csl: "/Users/theloniousgoerz/Academic/Professional/Resources/Citations/styles-master/american-sociological-association.csl"
editor_options:
  chunk_output_type: inline
  markdown: 
    wrap: sentence
geometry: "left=1in,right=1in,top=1in,bottom=1in"
number-sections: true
link-citations: yes
linkcolor: blue
font_size: 12pt
header-includes: 
  - \usepackage{framed} 
  - \usepackage{amsmath}
  - \usepackage{tcolorbox} 
  - \usepackage{pdflscape}
  - \newcommand{\blandscape}{\begin{landscape}}
  - \newcommand{\elandscape}{\end{landscape}}
  - \usepackage{multirow} 
  - \usepackage{xcolor} 
  - \usepackage{tabularray}
  - \def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}} 
  - \usepackage{setspace} 
  - \usepackage{threeparttable}
  - \usepackage{booktabs}
  - \usepackage{siunitx}
  - \usepackage{tikz}
  - \usepackage{tabularray}
  - \usepackage{float}
  - \usepackage{graphicx}
  - \usepackage{rotating}
  - \usepackage[normalem]{ulem}
  - \usepackage{csquotes}
  - \UseTblrLibrary{booktabs}
  - \UseTblrLibrary{siunitx}
  - \newcommand{\tinytableTabularrayUnderline}[1]{\underline{#1}}
  - \newcommand{\tinytableTabularrayStrikeout}[1]{\sout{#1}}
  - \NewTableCommand{\tinytableDefineColor}[3]{\definecolor{#1}{#2}{#3}}
  - \usepackage{bm}
  - \usepackage{endnotes}
bibliography: references.bib
---

```{r setup, include=TRUE, results="hide", echo = FALSE, message = F}

 rm(list = objects())
 # THIS IS THE WIDTH OF THE CODE                                                ###
 knitr::opts_chunk$set(echo = F,
                       resuls = T, message = F, warning = F)
 
 # knitr::opts_chunk$set(fig.width = unit(16,"cm"), fig.height = unit(9,"cm"))
 
```

```{r}
# Title page: 
# email: tlg95@cornell.edu
# Keywords: Segregation, Racial Inequality, Mortality, Demography
```

\singlespacing 

\section{Abstract}

Persistent Black-White segregation is a key mechanism shaping profound Black-White inequalities.
A central assumption is that Whites explicitly benefit from racial segregation and would lose advantages afforded to them if they were to residentially integrate with Black Americans.
The theoretical basis for this stems from a focus on neighborhood-level consequences of segregation that produce zero-sum outcomes.
However, segregation also shapes broader, macro-level dynamics, including income segregation, the availability of public goods, and policy outcomes that likely impact all residents.
In this paper, I track individuals born between 1905-1920 in the complete-count 1940 Census linked to Social Security death records and utilize an instrumental variable design that leverages county-level variation in government fragmentation to estimate the causal effect of segregation on Black and White longevity.
Results reveal that a 10-point increase in Black-White segregation decreased Black longevity by 0.44 years and White longevity by 0.29 years.
For Whites with lower education, segregation is particularly harmful, but for those with the highest education, the effects are attenuated by 0.10 years.
These results highlight that processes of racial stratification also affect Whites and contest straightforward zero-sum conceptions of segregation effects.

\pagebreak

\doublespacing
\fontsize{12}{22}
\setstretch{2}
\baselineskip=29pt

\section{Introduction}

```{r, echo = F, results = F}
rmdwc::rmdcountAddin()
```

The persistent segregation of Black Americans [@elbers2021] is both a mechanism of structural racism [@brown2024; @massey2003] and a fundamental dimension of Black-White inequality [@blokland2008].
A central assumption in this scholarship is that Whites benefit from racial segregation and will lose advantages afforded to them if they were to residentially integrate with Black Americans.
Zero-sum views about Black-White relations are pervasive and form the basis of Whites’ [@wetts2018; @warren2025] public opinions about redistributive policy [@gilens1996] and racial integration [@bobo1996].
Accordingly, Whites have continually resisted residential and educational integration due to a belief that segregation produces White advantage [@mcghee2021; @crowder2008; @rich; @logan2017; @johnson2019; @quillian2014; @frey1979].
In contrast, @dubois1935 concept of the “wages of Whiteness” casts doubt on this assumption, instead suggesting that zero-sum views—and racial divisions more broadly—serve the symbolic purpose of offsetting the effects of racial stratification on working-class Whites (e.g., through perceptions of higher status).

The underlying mechanisms of racial segregation are often described as occurring through the aggregation of micro-level processes that produce zero-sum outcomes [@quillian2014]; but this overlooks the fact that individuals spend substantial time outside their neighborhood, which expands their contexts of exposure.
Segregation also shapes macro-level outcomes that impact all residents, such as income segregation [@reardon2011], public spending on infrastructure and public goods [@chyn2023], and the political efficacy of policies supporting low and middle-income families [@ananat2009].
Because of the scope for neighborhood spillovers and the macro-level consequences of living in a segregated place, whether racial segregation actually leads to White advantages remains less clear.

In this paper, I test this assumption by examining the effects of racial segregation on the longevity of Black and White Americans using data from early 20th Century birth cohorts (1905-1920) in the 1940 Full Count Census linked to Social Security Numident death records (N = 2.1 million).
A fundamental challenge in determining the impacts of segregation is the endogeneity of place-level confounders like public policies and infrastructure that impact spatial sorting across places and ultimately affect health outcomes [@montez2020].
To address this, I utilize a validated instrumental variable approach that leverages variation in local government fragmentation, as measured by the number of local governments, within U.S. counties in 1967 [@cutler1997].\footnote{It is noted in (Cutler and Glaeser 1997) and Light and Thomas (2019) that the spatial arrangement of governments is largely constant over time and fixed well before the contemporary period and largely do not change over time. This gives additional confidence that contemporary segregation patterns do not shape government location and the degree of government fragmentation within places is not determined by other contemporaneous political or economic forces that may shape individual outcomes. This point is discussed further in the Methods section.} The intuition of this approach is that counties with greater government fragmentation in 1967 experience greater sorting between neighborhoods that increases segregation but is unrelated to individual outcomes.

Results reveal that a 10-percentage point increase in segregation decreased Black longevity by 0.44 years and White longevity by 0.29 years.
For Whites, the effects of segregation are particularly harmful for those with lower education; however, the highest educated White American’s longevity is reduced by 0.19 years.
These findings contest zero-sum conceptions of Black-White health inequality and highlight how processes of racial stratification to affect Whites by virtue of their class position.
These results suggest that policies promoting racial integration have the potential to improve both Black and White outcomes, rather than reduce White advantage.

\section{Background}
\subsection{Historical Processes of Racial Stratification and Zero-Sum Views Of Racial Inequality}

Although mechanisms of racial stratification such as political disenfranchisement and legal subjugation were targeted at Black Americans, historical record shows that Slavery and Jim Crow laws had significant scope to harm poor Whites as well.
As early as the 1850s, observers of chattel slavery in the Antebellum South recognized that processes of systemic racial inequality had profound impacts that were not confined exclusively to Black Americans.
In The Impending Crisis in the South: How to Meet, Hinton Rowan Helper outlined an argument for the abolition of slavery, in part because of its economic inefficiencies and disenfranchisement of poor, non-slaveholding Whites.
Helper demonstrated that free, Northern states had surpassed the South in trade, production, and economic diversity [@helper1857 :26].
Though slavery explicitly benefited elite, White Southerners, these advantages did not necessarily cascade down to poor Whites in the same way that Northern economic prosperity did.
Rather, Slavery resulted in widespread joblessness, economic insecurity, poverty, and illiteracy [@ash1991] among Black Americans, as well as poor Whites [@helper1857 : 377].
\footnote{The economic prosperity of the North is apparent in its direct effects on individual wages. For example, cotton factories in Massachusetts paid workers 80 cents per day compared to 50 cents in Tennessee (Helper 1857: 379), indicating large wage differentials between regions.} The scarcity of jobs and resources also reduced Whites’ ability to secure competitive wages in the labor market [@helper1857: 380] and played a part in wage depression more generally.

Following the abolition of slavery, efforts to undermine the economic and political incorporation of free Black Americans, in the form of Jim Crow laws, had similar implications for working class Whites.
Although it is well understood that Jim Crow laws ensured legal residential segregation, political disenfranchisement, and the policing of the everyday lives of Black Southerners, the methods employed to achieve these goals also implicated many poor White Americans [@perman2001 :2; @valelly2004 :134].
Despite a contemporary understanding that the goal of Jim Crow laws was to ensure White supremacy, historical evidence questions this interpretation [@perman2001 :6].
Jim Crow laws represented efforts by elite Whites to consolidate political power through voting rights restrictions, such as poll taxes, literacy tests, and arbitrary ballot procedures [@perman2001] as well as the intentional fueling of racial tensions between lower-class White and Black populations [@gabriel2023] .
Practically, these restrictions directly excluded Black and poor White populations from political participation, serving to fracture potential multiracial political coalitions [@perman2001 :2].

An illustrative example of these bi-racial impacts is evident in late 19th Century voting laws that sought to target Black voters, but in practice also suppressed poor White votes.
For example, the Dortch Bill of 1889, a set of voting restrictions in Tennessee, implemented secret ballots in cities with large Black populations and banned assistance at voting booths for illiterate voters [@turney1891] \footnote{The Dortch Bill, introduced by Tennessee Democrats in 1889, introduced a secret ballot meant to disadvantage illiterate voters by restricting assistance to illiterate voters. The sole purpose was to  “benefit the rich man, burden the poor one and in the end result in the defeat of the Democratic party in the State,’ (Perman 2001: 54). Quoting the Southwestern Christian Advocate (1891), Perman (2001) explains that in practice, this bill disenfranchised swaths of illiterate White voters. At the time, some Southern Democrats also objected to the Bill because of this loophole.The Dortch Bill was eventually ruled unconstitutional in Cook v. State of Tennessee (1891)}.
These measures restricted both illiterate Black voters and, unintentionally, large populations of illiterate Whites.
This ultimately led to significant reductions in votes for both Democrats and Republican candidates in the South [@perman2001: 54,59].
In both Antebellum and Jim Crow periods, historical evidence indicates that, though not the modal White experience [@ash1991], efforts and processes of racial stratification had scope for economic disadvantage and political disenfranchisement of Whites by virtue of their class position.

Despite evidence of the adverse effects of racial stratification on poor Whites, zero-sum views of race relations and racially coded issues [@gilens1996] are pervasive, far-reaching, and have lasting impacts on social policy and political movements [@wetts2018; @mcghee2021].
Public opinion research shows that White Americans believe that they are the beneficiaries of systemic racism [@quillian1995; @bobo1996; @quillian1996] and, by extension, view progress toward racial equality as coming at the expense of themselves.
Black Americans also endorse similar zero-sum views of Black-White race relations, which starkly contrast coalition-oriented relationships observed among other racial-group pairings and emphasize the exceptional nature of Black-White relations [@warren2025].
Explanations for why Whites hold zero-sum views are traced to perceptions of out-group threats to Whites’ advantaged position in society [@wetts2018; @blumer1958].
The presence of out-group threats, whether real or perceived, constitute challenges to their position [@blumer1958] and provoke responses of racial resentment [@gilens1996].
These beliefs then have significant consequences for political efforts and support for non-racial redistributive policies that are characterized through a racial lens [@gilens2000; @gilens1996].

At the core of Whites’ pervasive zero-sum beliefs about Black-White inequality is the assumption that processes of structural racism primarily advantage them.
Zero-sum views can be understood as part of a larger set of compensatory and symbolic mechanisms that offset the consequences of racial stratification on working-class Whites.
In Black Reconstruction in America, @dubois1935 provides an illustrative framework that reconciles Whites’ zero-sum beliefs with the deleterious consequences of racial stratification on working-class Whites.
In the context of the post-Reconstruction U.S., DuBois details how White political elites forged a powerful alliance with the White working-class by promoting the salience and superiority of White identity that served to counteract potential Black-White class solidarity [@dubois1935: 700] by casting \textit{all} Whites as the beneficiaries of racial stratification.

\begin{displayquote}
“It must be remembered that the [W]hite group of laborers, while they received a low wage, were compensated in part by a sort of public and psychological wage. They were given public deference and tides of courtesy because they were [W]hite,” (DuBois 1935: 700). \footnote{"Whiteness” is intentionally capitalized in quotations where it was originally lowercase to correspond with the capitalization conventions of the rest of the manuscript.}. 
\end{displayquote}

Meanwhile, Black Americans were subject to mob violence, racial terror, and public insult.
However, rather than racial subjugation resulting in the economic betterment of Whites, DuBois argues that the psychological wage paid to poor Whites, “was [such] that the wages of both classes could be kept low, the [W]hites fearing to be supplanted by Negro labor, the negros always being threatened by the substitution of [W]hite labor,” [@dubois1935: 701].
Within DuBois’ framework, zero-sum beliefs and racial divisions more broadly serve the symbolic purpose of overshadowing the shared class interests of working-class White, and Black populations by animating racial tensions.

In contemporary perspective, there are many parallels between racial segregation and earlier forms of racial stratification.
Residential segregation is widely understood as the "'structural linchpin' of race relations in America," [@bobo1996: 883] that links racial stratification to unequal outcomes [@dubois1899; @blokland2008; @light2019].
Historically, Black-White segregation has been slow to decline and in the wake of Civil Rights legislation outlawing housing discrimination and legal segregation.
Contemporary patterns of racial segregation are shaped, in part, by individual preferences and beliefs about racial integration and its consequences [@louie2024].
Despite notable progress toward in the past several decades, a stubborn barrier to desegregation is White resistance to residential integration through White flight [@crowder2008; @logan2017; @frey1979], avoidance from cities [@rich], and suburbanization [@owens2023].
However, empirical evidence for the assumption that racial segregation leads to White advantage remains limited.

\section{Theoretical Mechanisms Linking Racial Segregation to Longevity}

To evaluate the extent to which segregation disadvantages Blacks and advantages Whites, I study longevity as a powerful indicator of individual life chances [@weber1922].
Longevity is both an important dimension of population health as well as an important axis of social inequality [@wrigley-field2024; @maralani2021], including far-reaching implications for individuals’ social networks, collective memory, culture, political participation, and social movements [@wrigley-field2025].
Population health disparities between Black and White Americans are stark, in both contemporary and historical settings, and a key object of both scholarly and policy discussion.
For example, in 2023, the difference in life expectancy at birth between White and Black Americans was of 4.4 years [@arias2020].
Longevity thus comprises more than a single health measure and reflects the cumulative summation of health inputs over the life course.
There are a variety of contemporary mechanisms through which segregation may may affect longevity that unfold over the life course.
While most segregation scholarship focuses on the downstream consequences of the stratified neighborhoods that Black and White individuals live in, segregation also affects county-level processes that affect all residents.
In the following section, I demonstrate how (1) Neighborhood Effects and Spillovers, (2) Income Segregation, and (3) Political Context and Government Spending on public goods have significant potential to shape both Black and White longevity.
In the following section, I outline these mechanisms, but formal tests are beyond the scope of the paper.

\subsection{Neighborhood Effects and Spillovers}

A leading pathway for segregation to affect longevity is through the differentiation of Black and White Americans’ neighborhoods, leading to an uneven distribution of opportunities and risks.
The neighborhood environments that Black American live in tend to have higher poverty [@massey2003], elevated levels of crime [@light2019], greater exposure to hazardous conditions and pollutants [@wodtke2022], and lower quality educational opportunities [@wodtke2023; @goldsmith2009] than the neighborhoods of White Americans.
@massey2003 argued that increases in racial segregation create high concentrations of poverty, particularly for Black populations but also for Whites at smaller magnitudes [@massey1990].
However, @ananat2011 shows that segregation increased Black poverty while simultaneously reducing White poverty.
Poverty also indirectly affects health through declines in employment opportunities [@wilson1987] that strain individual resources [@aber1997].
The timing and duration of poverty are also consequential particularly as they relate to early life adolescence [@brooks-gunn1997; @sharkey2011; @sharkey2013; @chetty2016].
Further, @massey2018 found that neighborhood poverty was associated with accelerated aging among Black and White mothers, indicating the compounding effects of exposure over the life course [@diprete2006].

Another leading differentiator of Black and White neighborhoods is the concentration of crime and policing.
Past work has established associations between segregation, homicide victimization, and aggressive policing [@bell2020; @light2019].
Exposure crime is associated with worse health outcomes [@foell2021] and aggressive policing elevates risk of homicide [@edwards2019] in racialized ways.
Empirical evidence shows that segregation increases Black homicide risk and reduces Whites’ risk [@light2019].
Therefore, the concentration of crime in predominately Black neighborhoods serves as a buffer reducing exposure in predominately White neighborhoods.
Through this pathway, theoretical mechanisms and empirical evidence indicate that the consequences of segregation for crime exposure and their associated implications for health outcomes are racialized.

The residential separation of Black and White Americans can also affect resources and financial circumstances by increasing the cost-burden of housing in segregated places.
Because income and socioeconomic status are foundational drivers of health status, housing-related cost burden could have substantial effects on longevity albeit through differing pathways for Black and White populations.
@daniels1975 found that White renters, intent on living in predominately White neighborhoods, paid a premium to segregate from Black populations.
Meanwhile, @akbar2025 showed that Black homeowners in pre-war U.S. cities paid a 28 percent premium to purchase a home in a majority White neighborhood arising out of a combination of cost and stigma.
Housing cost burden could significantly affect longevity through reductions in spending power and potentially wealth accumulation [@hurd2003].

These neighborhood-level disparities contend a zero-sum relationship between segregation and health outcomes between Black and White Americans; wherein Black neighborhoods are exposed to higher disadvantage and White neighborhoods are shielded from it.
However, the emphasis on neighborhood effects often overlooks the dynamic environments that individuals encounter through everyday, routine exposures.
The variegated contexts that individuals navigate outside their neighborhoods provide significant scope for exposure to advantage and disadvantage that are not directly tied to residential neighborhoods [@browning2021; @cagney2020].
Individual daily mobility patterns show that, rather than being isolated within residential neighborhoods, urban populations are mobile throughout the day.
In any given day, individuals spend substantial amounts of time commuting to and from and at work and school, which expand the types of spaces they are exposed to.
@browning2021 estimated that high school-aged youths spend about 34 percent of their waking time outside their residential neighborhoods.
At the neighborhood-level, daily mobility flows are substantial, with inter-day mobility between neighborhoods being sufficient to radically reshape the demographics [@hess2024].
Given the profound role that non-neighborhood exposures play in everyday life, evidence indicates that activity spaces can expand individuals’ exposure to poverty, crime, and disadvantage also shape health beyond ones’ neighborhood.
@sharp2015 found that the interaction between neighborhood and activity space context affects health: individuals of disadvantaged neighborhoods reported better health when they spent time in more advantaged neighborhoods and vice versa.

Activity space exposures also shape the dynamics of cross-group contact which matter for mental health and wellbeing [@blumer1958].
While interpersonal contact between Black and White individuals is reduced by segregation, both neighborhood isolation and interracial contact through non-neighborhood exposures can affect health.
For example, @louie2024 argue that Whites’ isolation from Blacks lead to negative racial stereotypes that shape assessments of neighborhood safety, danger, and crime.
These negative stereotypes can lead to worse psychosocial distress and poor mental health for some Whites.
Moreover, decreased contact also shapes the experiences of Black residents when they navigate predominately White spaces.
@anderson2022 describes a parallel set of processes, where Black individuals visiting White neighborhoods are subject to greater monitoring, scrutiny, stigma, and microaggressions from Whites, which have similar implications for mental health.
Empirically, @hicken2014 found that Black Americans residing in Chicago reported higher levels of racism-related vigilance compared with Whites, which served as a mechanism for increased risk of chronic stress and hypertension.

\subsection{Income Segregation}

Another pathway, in addition to neighborhood-level mechanisms, are macro-level features of segregated places that shape the population health outcomes of all residents.
One mechanism that is particularly consequential for individual outcomes, such as economic mobility, is income segregation [@chetty2014 :34].
Because income and race are tightly correlated [@reardon2011; @massey1990] racial segregation is associated with income segregation, particularly among Whites [@jargowsky1996].

A significant body of literature has drawn links between income segregation and downstream effects on long-run individual economic and educational outcomes that shape health.
@chetty2014 found that income segregation significantly reduced the upward mobility of both Black and low-income White children (34).
Moreover, @chyn2023 also found that racial segregation also decreased the upward mobility of Black and low-income White children.
These effects on upward mobility are particularly consequential considering extensive research demonstrating the health returns to income and educational attainment [@montez2024].
Empirical evidence indicates that education increases longevity through a variety of economic and behavioral mechanisms [@halpern-manners2020; @lleras-muney2022; @montez2024].
For instance, @halpern-manners2020 found that an additional year of education resulted in .37 additional years of longevity for men born in the early 20th Century (1526).
Moreover, @finch2003 found that household income and material conditions were some of the most consequential determinants of infant health outcomes.
A more limited body of work has also found direct links between income segregation and health.
@mayer2005 found that economic segregation was associated with greater infant mortality.
Similarly, @ross2001 found that income segregation was associated with higher working-age mortality.

\subsection{Political Context and Government Spending}

Segregation also may impact longevity through its influence on political context, government infrastructure, and portfolios of public policies across places [@montez2020].
Prior literature has established that segregation is associated with lower spending on welfare, infrastructure, and public services [@chyn2023; @ananat2009].
Administrative units, such as the county or state, are important for health because they are intrinsically linked to important political, public health, legal, and institutional processes that affect individuals living within those boundaries.
Places with more politically conservative policy portfolios regarding reproductive health, spending on education, income support, labor protections, and housing support [@montez2020: 672] were associated with reduced life expectancy among both men and women [@montez2020: 668].
More generally, ethnographic work has provided support for these patterns, showing that low-income Whites, especially in rural areas, are particularly vulnerable to the health effects of conservative policy portfolios [@metzl2019].

Beyond policy decisions that concern health, racial segregation also shapes public support and efficacy of progressive policies related to employment, labor relations, and redistributive policy.
@ananat2009 found that segregation reduced the efficacy of working-class oriented policies that would improve the economic position of low-income residents.
@chyn2023 also found that racial segregation produced higher levels of racial resentment among White Americans which may serve as a mechanism to decrease electoral support for progressive policy.
Whites’ racialized attitudes have long been shown to have implications for views about welfar and redistributive policy, which have direct implications for electoral outcomes [@wetts2018; @gilens1996].
Indeed, @chyn2023 also found that racial segregation was associated with decreases in spending on social welfare, public goods, and infrastructure.
Decreases in funding for social welfare programs providing income and nutritional support have particularly strong implications for the health and wellbeing of low-income individuals reliant on these services regardless of race.

Moreover, a key dimension of government infrastructure is funding for public health services that have contributed to significant contemporary and historical improvements in population health [@cutler2005; @liang2023].
Reductions in spending on public health services and infrastructure thus have important consequences for population health and disparities more generally, regardless of residence within an area.
Through this mechanism, variation in county-level health services and infrastructure could significantly shape longevity.

\section{Prior Work on Racial Segregation and Health}

Past work has established that racial segregation is associated with negative health outcomes for Black Americans and wide Black-White health disparities.
While most studies focus on the health outcomes of Black Americans, a limited number of studies consider the effects of segregation on White Americans, but with mixed results due to methodological limitations.

One strand of research has drawn links between racial segregation and infant birthweight.
@austin2016 found that segregation decreased the birthweight of Black children by 2.8 grams and decreased the birthweight of White children by 0.63 grams.
However, another study by @vu2024 found negative impacts of segregation on the birthweight of Black children and no effect on the birthweight of White children.

On mortality, @logan2018 found that, in certain contexts, segregation was associated with greater mortality risk for White Americans and lower mortality risk for Black Americans.
Similarly, @karbeah2023 found that segregation was associated with a greater risk of mortality among White and Black children.
However, the generalizability of the conclusions from @logan2018 are limited because they rely exclusively on death records from North Carolina.
Moreover, @karbeah2023 use a measure of segregation that is incongruent with established segregation theory and measurement, because of data limitations.
Instead, their measure reflects processes at the neighborhood-level, because their measure of segregation is based on the racial composition of one's neighbors, rather than the distribution of groups within a place.
Because this measure reflects a different set of processes, which the authors acknowledge, than what is conceptualized by contemporary segregation scholarship, it is unclear how to understand these results in the context of established measures of segregation.

While existing studies provide evidence of associations between segregation and poor health outcomes, particularly for Black Americans, results across studies for White Americans are mixed and are largely correlational.
Demographic work has demonstrated that there are many aspects of places, such as the policy context, infrastructure, and environment, that are consequential for population health that are related to segregation [@montez2020].
Currently, these results provide suggestive evidence but are not able to establish causality.
Therefore, if residential segregation and longevity are shaped by unobserved heterogeneity or confounded by other features of counties, then correlational analyses cannot distinguish whether relationships are causal.

Finally, it is important to distinguish the relationship between segregation and health and health disparities.
Using life table methods, @hendi2024 showed that segregation was responsible for a 16-year increase in the Black-White life expectancy gap for men and a 5-year increase for women between 1990-2019.
While studies of Black-White longevity gaps are informative for characterizing population-level inequalities, they obscure important information about the direction of these effects and are substantively ambiguous without further context [@wrigley-field2025].
If segregation positively affects White health and Black health negatively, or affects both groups negatively, then these relationships both produce inequality.
Distinguishing between these relationships is important, because longevity inequalities that arise from the same or different signed relationships point toward distinct implications for sociological theory.

\subsection{The Present Study}

The goal of this paper is to estimate the plausibly causal effect of segregation on the longevity of Black and White Americans born to early 20th Century birth cohorts and residing in U.S. counties between 1980-2000 in later life.
To establish the effects of segregation on longevity, I rely on a validated instrumental variable approach that leverages variation in local government fragmentation within U.S. counties in 1967.
This instrument induces quasi-random variation in racial segregation, through a Tiebout mechanism.
A Tiebout mechanism is the tendency of individuals to "vote with their feet".
In areas with more local governments, there is a greater variation in residential options in the form of differing tax rates and services, which allows individuals to sort according to their preferences.
Greater sorting leads to higher segregation because households with similar backgrounds and resources cluster in similar areas, but does not necessarily affect individual outcomes.
I combine this quasi-random variation with individual-level full count Census data linked to Social Security Numident mortality records.
In accordance with historical record and prior scholarship, I also examine how the effects of segregation on longevity vary by educational status to distinguish whether educational status moderates the effects of segregation on longevity.
Establishing the causal effects of segregation on Black and White longevity contributes important clarity about whether Whites' zero-sum beliefs reflect realized benefits from segregation or symbolic benefits that serve as a mechanism to offset the effects of racial stratification.

\section{Data}
\subsection{Sample and 1940 Linked-Census Data}

The empirical analysis of this paper is anchored by individual-level demographic data records from the 1940 full fount Census [@ruggles].
To construct my analytic sample, I identify individuals from the 1905-1920 birth cohorts who were 20-35 in 1940.
These individuals are then linked to Social Security Administration (SSA) mortality records between 1988 and 2005 provided by the Berkeley CenSoc Numident file [@breen2023].
These data offer unparalleled individual-level data on mortality that are geolocated to contemporary residence at death and which are representative of the population living in the United States in 1940.
CenSoc Numident records cover deaths occurring between 1988 and 2005, allowing for reliable estimates of longevity conditional on an individual surviving to age 65 [@goldstein2023]\footnote{The interpretation of longevity conditional on surviving to 65 is an artifact of using truncated mortality data which only reliably cover deaths over the age of 65 (Breen et al. 2023: 35) and is interpretation used in (Breen 2024).}.
The geographic specificity and rich covariates make these data ideal for analyzing mortality and longevity disparities.

I focus on the 1905-1920 birth cohorts for two reasons.
First, their birth years and the end year of the mortality observation window allow for plausibly complete observation of mortality that is not biased by right censorship (i.e. missing the oldest old).
In other words, those born in 1905 would be 100 years old in 2005, making it reasonable to assume that few individuals remain alive.
Even for those individuals who live very long, using doubly truncated data will result in estimates biased toward zero [@breen2024], and thus produce estimates of the effect of segregation on longevity that are likely conservative.
Second, restricting to ages 20-35\footnote{Halpern-Manners et al. (2020) finds that the mean number of years of education of decedents in the 1940 Census is 10.4 years (1525). If we assume a normative time of starting school at 6, individuals in the sample reach the average at 16.4 years, which is well before our age cutoff.} allows for confidence that individual-level confounders to longevity, such as education status, are observed.
By conditioning on those 20 years or older, it is reasonable to assume that individuals have completed the modal level of formal education by the time that they appear in the sample.

It is well known that the Numident data suffer from the problem of double truncation, where data are truncated on both the left and right side.
@goldstein2023 show that when not accounted for, doubly truncated mortality observation windows result in biased estimates of longevity disparities toward 0 and higher mean observed death age in samples.
This means that those born in 1905 are observed between the ages of 83-100 years old and those born in 1920 are observed between the ages of 68-85 years old.
Birth year fixed effects are integral to reducing bias in mortality analysis with truncated data [@goldstein2023], where individuals of varying birth cohorts may have differing numbers of years covered by the death window.

There are also representativeness features of the data that bear on substantive interpretations.
The CenSoc Numident file links 1940 Census data to the public release of the National Archives Record Administration “NARA Numident” file using the Abarmitzky, Boustan, and Eriksson (ABE) exact matching algorithm [@abramitzky2012].
For Black Men and Women, comparisons with the Census indicate that the Numident file slightly underrepresents these groups [@breen2023].\footnote{For Black Men the Numident links are slightly underrepresented at a -1.8 percent difference from the Census and slightly more at a -2.4 percent difference (Breen et al. 2023: 8-9). This matching procedure may be particularly selective for unmarried women, because of changes of marriage-related name changes over the life course. Fortunately the Numident data contain parents last names that can be used to link women who have undergone name change.} To add further validity to these estimates and address representation issues, I use post-stratification weights developed by @osborne that re-weight CenSoc Numident counts to period-specific distributions of deaths to adjust for possible selection.
Table 1 describes the representativeness of the population.

\subsection{County-level Data}

I examine exposure to segregation in mid-to-late life, obtained and calculated based on the county individuals resided in at death.
Examining the county at death offers an empirical understanding of the effects of segregation that occurred after the Great Migration and the passage of Civil Rights Act that outlawed legal segregation.
Because much of the contemporary segregation scholarship examines the consequences of de jure segregation, this time period aligns more consistently with these theories.
Scholars recognize that the geographic scope and structure of segregation after the Great Migration are understood to be distinct from the segregation that occurred before, predominately in the South [@leibbrand2020; @grigoryeva2015].

Despite potential limitations of using one’s county at death, it is reasonable to believe that place characteristics are highly correlated with their residential experiences at birth.
Patterns of neighborhood attainment and residential mobility over the life course show that the modal neighborhood transitions are to places with similar characteristics [@sampson2008].
Moreover, across the life-course, migration propensity declines with age [@bernard2014] with individuals less likely to migrate after age 40 old and, when they do, it is usually short distances [@molloy2011; @zimran2022].
For these reasons, segregation in the county of death likely represents a period of exposure encompassing mid-life and is correlated with earlier experiences.
However, to account for any residual variation due to early life exposures and the exceptional nature of moves during the Great Migration, I include birth county fixed effects in all models.

To measure segregation, I use the dissimilarity index (D) [@duncan1955].
The dissimilarity index measures the evenness of the distribution of Black and White residents within a county, ranging from 0-100, and is interpreted as the share of Black or White individuals that would have to move to a different neighborhood for the racial composition of neighborhoods to be the same as the racial composition of the county.
The dissimilarity index is widely used in prior work [@hendi2024; @light2019; @massey2003; @cutler1997] because it is straightforward to calculate, concordant with past empirical evidence, and has a concrete interpretation.
Nevertheless, the D is well known to suffer from various index biases, outlined in @fossett2017 and others [@winship1977; @carrington1997], particularly as they relate to temporal comparisons and variability in the population composition and relative group size of counties.
Accordingly, in supplemental analyses, I replicate the main results with a bias-adjusted D index that adjusts for this bias [@fossett2017]\footnote{In contrast to D, the bias-adjusted D measure is more complex to estimate and the interpretation less concrete (Fossett 2017: 233). For these reasons, I use the conventional dissimilarity index for main analyses.}.The dissimilarity index for the county an individual resided in at the time of their death is calculated formally in Equation (1):

\begin{equation} 
D = \frac 1 2 \sum_i^N | \frac{b_i}{B} - \frac{w_i}{W}| \times 100 
\end{equation}

Where N is the number of counties, $b_i$ and $w_i$ are the proportion of Black and White residents in the $i$th census tract and B and W are the proportion of Black and White residents in the county.
The dissimilarity scores are calculated from decennial Censuses 1980, 1990 and 2000, which correspond to the year of death observed for sample members.

Counties are additionally linked to 1967 Census of Governments data on the number of municipal and township governments and their federal revenue share from the Government Finance Database [@pierson2015] that I use for identification.
My final analytic sample consists of ($N = 2,094,017$) individuals residing in $1,774$ counties at their time of death over three decades from 1980-2000 (Table A1).
The resulting counties are filtered from all counties in the U.S. such that they have a Black population greater than 0 and are not missing data in 1967 on the county finance measures used as instruments \footnote{The data filtering procedure and stepwise sample sizes are included in table A1 in the Appendix. Beyond filtering to counties that have non-zero Black populations, the sample size is further reduced by variability across years in Census of Government reporting rates. Pierson et al. (2015) explain that annual sample sizes of governments vary widely, which result in missing data, and reporting at smaller government levels, such as the township and municipality, vary in size and quality year-to-year (22). However, years ending in 2 and 7 are the most representative, making 1967 an ideal candidate year because it is the earliest year that data is available and of the highest relative quality in terms of data coverage (Pierson et al. 2015 :22).}.Table 1 displays the individual and county characteristics between those in the 1905-1920 birth cohorts in and out of the sample and demonstrates that the differences are minimal.

\section{Methods}
\subsection{Empirical Strategy}

There are many features of counties and places that also shape health [@montez2020], leading associations between county-level segregation and longevity to be biased.
A leading strategy employed in prior scholarship to address concerns of endogeneity are instrumental variables that create exogenous variation in segregation but do not plausibly affect health.
I use a validated instrument first used in @cutler1997 and employed subsequently by @light2019 that leverages variation in the levels of government fragmentation of counties in 1967 that induce geographic sorting between neighborhoods.
I instrument for segregation using (1) the number of municipal and township governments within a county and (2) the share of state intergovernmental revenue allocated to these local governments.
The intuition behind this approach is that areas with more local governments offer residents more opportunities to “vote with their feet” in selecting among jurisdictions that differ in taxes, services, and demographics.
This greater potential for sorting fosters more residential segregation, as households with similar preferences and resources cluster within the same localities.

The use of municipal and township governments to instrument for contemporary population patterns are appealing for multiple reasons.
First, as demonstrated by @cutler1997 and @light2019, the number of governments over time is largely constant, with a correlation over time of $.98$ [@cutler1997 :848].
Certain government patterns are fixed stretching as far back to 1787, suggesting that government placement is not affected by contemporary sorting patterns.
Second, because of their constant nature, the number of governments are likely to result in spatial sorting that is informed by quasi-random early settlement geographic features, such as the arrangement of hills, rivers, and other geologic features that affect segregation patterns but are unrelated to contemporary demographic forces that affest segrgation and health outcomes [@cutler1997].
An implication of this constant nature is that, as segregation has declined over time, the predictive association between governments and segregation should also decrease over time.
I consider the impact of these over-time changes in the discussion of IV assumptions.

In analyses, I include the log of the number of governments and the state-level share of intergovernmental revenue to purge local endogenous factors.
When evaluating the plausibility and relevance of an instrument, it is important to do so in the context of IV assumptions [@felton2024]: (1) the instrument is exogenous and relevant; (2) there is a strong first stage relationship; and (3) monotonicity or the instrument only influences the level of segregation in a county in one direction.

The first assumption is that these instruments impact longevity only through segregation.
To further validate this assumption, I follow @cutler1997 and adjust for inter-governmental revenue shares which have the advantage of accounting for endogenous variation in resources at the local government level that may also affect longevity \footnote{Because there are more instruments than endogenous regressors, a natural test of the validity of many instruments are overidentification tests. These tests are commonly interpreted as testing the” validity” of the instruments, however overidentification tests must first assume that one instrument is valid (e.g., in the just-identified case) (Felton and Stewart 2024), which is not informative for assessing validity. A more informative interpretation from Mogstad et al. (2021), is that rejection of overidentification tests suggest that there is heterogeneity in treatment effects across subpopulations affected by the different instruments, which is not in reference to validity. Additionally, the overidentification tests assume constant treatment effects, which is unreasonable in a setting where the IV estimate is the local average treatment effect. Using the same instruments, Cutler and Glaeser (1997:854) fail to pass overidentification tests for the instruments because of large sample sizes ranging from 3-6 thousand observations. As expected, the instruments generally fail to pass these tests due to the large sample size of my data where observations are in the millions. An alternative to overidentification tests is placebo analyses that require that a researcher find an additional instrument that shares all the same confounders as the original instrument but is thought to not affect the treatment (Felton and Stewart 2024). While informative, the shared confounders assumption of placebo instruments is also quite strong and the search for a candidate placebo instrument faces the same challenges as credible instrument.}.

It is also important to recall that the instrument is the density of governments in a county and does not rely on any government-specific quality metrics.
Second, in Figure 1 (Table A2) I demonstrate that association between the instruments and segregation at the county-level is highly predictive using a binned scatter plot.
Tables of the main IV models also show the first stage for individual-level regressions.

\begin{center}
[Figure 1 here]
\end{center}

The first-stage F-statistic assesses the strength of the instrument and segregation and indicate a strong and significant F-test of $226.53$.
This test gives confidence in the assumption that governments strongly predict racial segregation and are not subject to bias that may result from weak instruments [@angrist2021].
Third, monotonicity requires that the instrument pushes segregation in the same direction for all units.
While a direct test of this is not possible, it is possible to examine patterns consistent with monotonicity to provide evidence in support of this assumption.
Accordingly, I employ a similar test to @rauscher2020 to examine demographic patterns that would be consistent with monotonicity.
It is well known that segregation levels have declined over time.
If we believe that part of this decline is due to less economic and racial residential sorting, then we would also expect that year-over-year the association between the instruments and segregation would become weaker.
One way to do this is to examine how the strength of the association between the instruments and segregation changes decade-over-decade.
Consistent with this assumption, the strength of the association between the instruments and segregation, as measured by $R^2$ decreases in size between 1980-2000.
While this does not represent a conclusive test of monotonicity, these patterns are consistent with the assumption.

Conditional on these assumptions, IV estimates of a regression of the effects of segregation on longevity are interpreted as the local average treatment effect (LATE) [@imbens1994; @mogstad2021], or the treatment effect amongst those living in counties where segregation increased because of greater government density.
These instruments have been used effectively in prior studies to examine how segregation effects multiple outcomes [@cutler1997; @light2019].

\subsection{Estimation}

Empirically, I estimate the relationship between segregation and age at death, conditional on living to age 65, in the county of residence at death.
I estimate the model via two-stage least squares estimation, stratified by racial group.
I control for a rich set of individual and contextual covariates to adjust for any residual confounding.
In fully specified models, I control for sex, education, marital status, employment status, whether an individual migrated, whether they reside in the South, urban-rural fixed effects \footnote{Between 1993 and 2003, Rural-Urban Continuum codes changed from their definitions between 1983-1993. For 1983-93 codes, metro counties were classified using a scheme ranging from 0-3, with counties that belonged to a metro area with a population of 1 million or more were categorized into Central and Fringe counties. In 2003, this scheme was revised so that counties were classified based only by population, with Fringe and Central collapsed into one category. Accordingly, to increase comparability. I collapse codes 0 and 1 in the 1983 and 1993 continuum code schemes into one category so that they are harmonized 1-3 scheme used in 2003.}, birth year and state fixed effects, and occupation fixed effects.
In all models, I include robust standard errors to account for heteroskedasticity.
Equation (2) describes the model\footnote{I also re-run models with standard errors clustered at the county-level the results are robust to alternative standard error specifications.}:

\begin{equation} 
Y_i = \beta_0 + \beta Segregation_{i} +  \pmb{\beta} X_i + \gamma_{b(i)} + \lambda_{c(i)} + \epsilon_i 
\end{equation}

The primary goal is to estimate $\beta segregation$, which is our focal measure of segregation operationalized as the dissimilarity in an individual’s county at death.
$\pmb{\beta}$ is a vector of regression coefficients and $X$ is a vector of individual-level control variables and $\gamma$ and $\lambda$ are fixed effects for individual birth year and birth county.

In summary, a core strength of this approach is that the instruments address critical endogeneity and selection challenges to estimating the relationship between segregation and longevity and rich individual-level covariates address additional confounding.
The Results section proceeds as follows.
The first part describes the characteristics of individuals in my sample and presenting results of the OLS association between segregation and longevity.
The second part presents the instrumental variable results, heterogeneity by educational status, and a set of robustness tests.

\section{Results}
\subsection{Main Results}

First, I present descriptive statistics of the analytic sample (N = 2.1 million) compared to the full sample in Table 1 and by racial group for the analytic sample in Table 2.
Between samples, the analytic sample appears to be broadly representative.
However, analytic sample members live in counties that have larger populations, are slightly less likely to live in the South, and are more educated than those in the full Numident sample.
By racial group (Table 2), White Americans have a clear longevity advantage over Black Americans and are more educated.
Black Americans are significantly more likely to live in the South and live in counties with a significantly higher Black populations that are more segregated than White Americans.

Next, I turn to the results from OLS regression models of the associations between segregation and longevity.
Figure 2 demonstrates that segregation is associated with reductions in the longevity of both Black and White Americans with Black Americans facing a substantial mortality penalty relative to White Americans.
A 10-point increase in dissimilarity is associated with a 0.84-year decrease in years of life for Black Americans and a 0.72-year decrease for White Americans (Table 3).
These associations are also robust to the inclusion of covariates and fixed effects in the adjusted models.

\begin{center}
[Figure 2 here]
\end{center}

Next, I turn to the results from instrumental variable regression models estimated via two-stage least squares.
Figure 3 presents instrumental variable estimates of the effect of segregation and longevity.
These estimates confirm the OLS finding that segregation reduced the longevity of Black Americans and White Americans but with a notable reduction in magnitude.
The unadjusted models indicate that a 10-point increase in segregation reduces Black longevity by 0.40 years and White longevity by 0.36 years.
Adjusting for covariates increases the magnitude of the effect of segregation for Black Americans to 0.44 years and attenuates the association for White Americans further to 0.29 years.
These results indicate that racial segregation substantially reduces the longevity of Black Americans and modestly of White Americans.
In comparison to OLS associations, IV results suggest that the OLS significantly overstates the magnitude of effects for Black and especially, White Americans.
The difference in the respective magnitudes of the OLS and IV coefficients is likely due to unobserved heterogeneity at the county and individual level.

\begin{center}
[Figure 3 here]
\end{center}

In addition to main estimates, I also stratify IV models by gender to understand the degree to which gender and race intersections may matter for the effects of segregation.
Figure 4 (Table 5) shows that the effects of segregation on longevity for Black men and women are virtually indistinguishable (diff = 0.02), but there is a slightly larger reduction in the harmful effects of segregation for White women’s longevity compared to White men.

\begin{center}
[Figure 4 here]
\end{center}

A 10-point increase in segregation corresponds to a 0.31-year decrease in longevity for White men compared to a 0.28-year decrease for women – a difference of 0.03 years.
In combination with the main results, results by gender indicate that there are inequalities among Whites by gender that advantage White women compare do men, however the mechanisms leading to this difference beyond well-known gender differences in mortality are beyond the scope of this paper.

\begin{center}
[Figure 5 here]
\end{center}

Next, I address a supporting research question that examines how the effect of racial segregation on longevity varies by educational status.
Recalling predictions from DuBois’s wages of Whiteness theory, if Whites are differentially harmed by racial stratification, then it would be expected that the effects of segregation on longevity vary by educational status.
Figure 5 and 6 (Table 6) display the results from the IV models of racial segregation on longevity interacted with education level for White and Black Americans.
First, results for Whites reveal that education is a significant moderator of the effects of segregation on longevity, such that the most advantaged Whites are harmed less by segregation.
For those with a college degree or higher, segregation reduced White longevity by 0.19 years.
In contrast, those with some college or less have the largest longevity penalties.
These results underscore a clear class gradient to these effects, suggesting that higher education buffers the negative effects of segregation on Whites.
In contrast, Figure 6 (Table 6) demonstrates a remarkably consistent effect of segregation for Black Americans, regardless of education.
Despite variability in the coefficients by education level, statistical tests of the interaction and differences between coefficients show that even at higher levels of education, segregation substantially reduces Black longevity.

\begin{center}
[Figure 6 here]
\end{center}

\subsection{Robustness Checks}

As a robustness exercise, I examine results using post-stratification weights to adjust for potential selection relative to period-specific death rates, and by using the Theil H index [@iceland2004], Isolation index, and bias-corrected Dissimilarity Index as alternative measures of segregation.

First, I re-estimate IV regressions using CenSoc post-stratification weights that adjust for lingering representativeness issues (Table A3).
Results from weighted regression models increase the magnitude of effects for both White and Black Americans with and without controls compared to unweighted results and underscore the same relationships as the main results.

Second, I examine the main IV results with two alternative measures of segregation: the entropy-based Theil index and the isolation index, which measure the probability that Black and White individuals interact within a geographic parcel (Table A4).
Using the isolation index, Theil H index, and bias-adjusted D, these alternative measures of Black-White segregation resoundingly confirm the main empirical results using the dissimilarity index.
For Black results specifically, the H index attenuates the estimated effect slightly, but the sign and significance of the effect is unchanged.
The bias-adjusted D also confirms these results.
Overall, across different subgroups, specifications, and measures, the results are quite consistent.
Racial segregation substantially decreases the longevity of Black Americans and modestly decreases the longevity of White Americans.

\section{Discussion and Conclusion}

Guided by zero-sum theories of processes of racial stratification, the prevailing expectation is that segregation positively increases White longevity and reduces Black longevity.
Though past work finds associations between racial segregation and Black-White health disparities, correlational studies are limited by endogeneity and unobserved heterogeneity that cannot be interpreted as causal.
Despite expectations that segregation unambiguously advantages Whites, historical record indicates there is significant scope for processes of racial stratification to have unintended effects on lower-class White Americans.
In this paper, I test the assumption that White Americans necessarily benefit from racial segregation using linked 1940 full count Census data and an empirical strategy leveraging variation in the government fragmentation of counties.

In naïve OLS regression models, IV models that address selection and endogeneity concerns, a variety of alternative specifications, and across alternative measures of segregation, the empirical results are highly consistent.
Contrary to zero-sum expectations, racial segregation substantially and significantly reduces Black longevity and modestly reduces White longevity.
It would be inaccurate to claim that residential segregation harms Black and White Americans in the same ways; rather, segregation reduced Black longevity at a magnitude approximately 50 percent greater of White Americans.
At the same time, these results reject straightforward zero-sum assumptions of White advantage.
Though modest, these effects suggest that desegregation may positively benefit White Americans, rather than adversely affecting their life chances.
In the context of noted declines in racial segregation between the 1970s to 2000s of approximately 21 points \footnote{The 21-point difference is derived from the author’s calculation the difference between the average levels of segregation in sample counties in 1970 and the average in 2000.}, a shift from 1970 to 2000 average county racial segregation corresponds to a gain of 11.09 months of life for Black residents and 7.31 months for White residents.
In the context of other studies concerning longevity, these results are substantial.
For instance, using the 1940 Full Count data, @breen2024 estimates that homeownership increased longevity by .31 years (or 4 months) and @halpern-manners2020 estimate that each \textit{additional} year of education increased longevity by .35 years (or 4.2 months).
These results speak to far-reaching impact of racial stratification for the lives of Black and White Americans [@louie2024; @mcghee2021].

In connection with the main results, it is important to emphasize that the effects of segregation are differentiated by educational status for White Americans but not for Black Americans.
For White Americans, attaining a college degree or higher is associated with modest reductions in the harmful effects of segregation on longevity, however this does not entirely offset longevity reductions.
Importantly, results overall contest zero-sum expectations, but reveal important heterogeneity in the effects of segregation on longevity, with highly educated Whites affected the least.
By contrast, for Black Americans, the effects of segregation are equally harmful regardless of education level.
The inequality in the protective effects of education by race is consistent with a growing literature on differential health returns [@deangelis2022] that finds that Black Americans, even at higher statuses, do not reap the same health benefits as similarly advantaged White peers.

This paper makes several important contributions to research on the consequences of segregation as well as the sociology of race and race relations.
Prevailing views on racial stratification, of which segregation is a key mechanism, have focused on how segregation creates divergent outcomes between Black and White residents with Blacks suffering the harmful ramifications and Whites reaping the benefits.
Two past studies, @light2019 and @quillian2014, test the hypothesis that Whites benefit from segregation in terms of homicide victimization and educational attainment that are outlined in terms of zero-sum neighborhood effect mechanisms.
While @light2019 finds support for a zero-sum relationship between segregation and crime, @quillian2014 finds that segregation decreases Black educational attainment but is ambiguous for Whites.
Despite the salience of the Black and White neighborhoods for outcomes, these perspectives do not fully capture the myriad mechanisms of segregation and overlook the role that place-level processes and variegated neighborhood exposures play.
The structural effects of segregation shape entire places and thus have the potential to shape experiences of all groups regardless of residence or relative neighborhood isolation.
Results in this paper emphasize that despite pervasive zero-sum expectations, racial segregation negatively shapes the population health outcomes of both Black and White Americans.
Overall, these results contribute to a growing literature on the generally muted effects of desegregation on White outcomes [@anstreicher2022; @johnson2019].

In concordance with theories developed by @dubois1935, the wages of Whiteness serve a compensatory role in offsetting the pernicious, class-based effects of racial stratification on lower-status White Americans.
Zero-sum expectations that Whites are better off in a racially stratified society and disadvantaged by desegragation are symbolic mechanisms to offset the harmful effects of racial segregation on low-status Whites by virtue of their class position.
The unequal effects of segregation by educational status show that only highly educated Whites benefit while less educated Whites do not.
Moreover, concordant with historical record, these results demonstrate that the educational and social class dimensions of processes of racial stratification affect Whites.
Like earlier efforts to disenfranchise Black Americans under Jim Crow, these efforts carry significant scope for class-based inequalities that transcend race.
The class-specific consequences of segregation highlight that in addition to reductions in Black and White longevity overall, Whites with greater education are better off but still harmed by racial segregation.

This study is also not without limitations.
First, there is inherent selectivity embedded in examining years of life conditional on surviving to 65.
While this analysis is informative for understanding longevity in this period, it cannot speak to inequalities that may arise in early and mid-life that may differentiate Black and White populations.
However, because these results reveal substantial effects of segregation on longevity even in this selective population, the disparities observed within the full population may be larger.
Moreover, the doubly-truncated nature of these data result in regression estimates that are biased toward zero, suggesting that these estimates are conservative.

Second, while instrumental variables are powerful causal inference tools to address selection in cross-sectional settings, they require strong assumptions that cannot directly be tested..
I demonstrate that the relevance condition is strongly supported and descriptive evidence consistent with monotonicity.
Relying on prior research and reasoning, the historical nature of the instruments situated well-before the study is suggestive that exclusion is supported, but the exclusion restriction cannot be directly tested.
Despite these limitations, and differences in the magnitudes of the coefficients, the conclusions from OLS and IV regression models are the same.

Third, while this analysis leverages historical birth cohorts, these results cannot speak to more contemporary relationships.
Because, segregation has declined over time, it is reasonable to believe that effects for more recent birth cohorts may be attenuated compared to this population.
Nevertheless, it is important for future research to consider how the effects of segregation on health have changed over time.

These limitations provide promising suggestions for future research.
Beyond using full count census data and information on completed mortality, future research would be to leverage representative surveys and information on health conditions that are predictive of mortality.
For instance, examining outcomes like Allostatic Load could allow researchers to gain a contemporary understanding of the ways that segregation shapes processes relevant to mortality.
While the study of mechanisms is beyond the scope of this study, I have outlined several compelling population and neighborhood-level mechanisms of segregation that plausibly affect individual outcomes.
Future work could consider studying these mechanisms directly to determine the primary mechanisms through which segregation affects individual outcomes and whether mechanisms vary by the outcome of interest.

If there is one takeaway from this paper, it is that the processes of racial stratification have severse consequences for the lives of Black and White Americans.
In the context of historically fraught Black–White relations in the United States, Whites’ persistent resistance to racial integration remains a barrier to desegregation.
A central component of this resistance is the misinformed belief that racial segregation necessarily advantages Whites.
Although zero-sum understandings of racial inequality are widespread, this study shows that segregation also harms Whites.
While those with greater education are buffered from the harmful effects of segregation, those advantages are modest.
For the vast majority of Whites, segregation negatively affects life chances by reducing longevity.
Crucially, this paper demonstrates that processes of racial stratification are also fundamentally processes of class stratification.
In line with @dubois1935, evidence of segregation’s biracial consequences lends credence to the view that casting Whites as unequivocal beneficiaries of segregation obscure its class-based harms which has implications for organizing and collective action rooted in shared circumstances.
Consistent with prior scholarship, the findings suggest that policies promoting racial integration could improve health and life chances for both groups while narrowing Black–White disparities.

\section{References}
\singlespacing

::: {#refs}
:::

\newpage

\section{Tables and Figures}

\graphicspath{ {../../bw_segregation_longevity/FigTab/}}
\makeatletter
\def\input@path{{../../bw_segregation_longevity/FigTab/}}
\makeatother
\begin{figure}[h!tbp]
\centering 
\includegraphics[width=.9\textwidth,height=.9\textheight,keepaspectratio]{fs_plot.jpeg}
\caption{First-Stage Relationship}
\end{figure}
\begin{figure}[h!tbp]
\centering 
\includegraphics[width=.9\textwidth,height=.9\textheight,keepaspectratio]{ols_figure.jpeg}
\caption{OLS Estimates of the Association Between Segregation and Longevity }
\end{figure}
\begin{figure}[h!tbp]
\centering 
\includegraphics[width=.9\textwidth,height=.9\textheight,keepaspectratio]{iv_plot.jpeg}
\caption{IV Estimates of the Effect of Segregation on Longevity }
\end{figure}
\begin{figure}[h!tbp]
\centering 
\includegraphics[width=.9\textwidth,height=.9\textheight,keepaspectratio]{iv_plot_gender.jpeg}
\caption{IV Estimates of the Effect of Segregation on Longevity (By Gender)}
\end{figure}
\begin{figure}[h!tbp]
\centering 
\includegraphics[width=.9\textwidth,height=.9\textheight,keepaspectratio]{White_Ed_Interaction.jpeg}
\caption{IV Estimates of the Effect of Segregation on Longevity By Education (White) }
\end{figure}
\begin{figure}[h!tbp]
\centering 
\includegraphics[width=.9\textwidth,height=.9\textheight,keepaspectratio]{Black_Ed_Interaction.jpeg}
\caption{IV Estimates of the Effect of Segregation on Longevity By Education (Black) }
\end{figure}
\input{representivity_table.tex}
\input{descriptive_table.tex}
\input{OLS_results_table.tex}
\blandscape
\input{IV_results_table.tex}
\elandscape
\input{iv_results_table_gender.tex}
\input{IV_by_Education_table.tex}

\section{Appendix}

\setcounter{table}{0}
\setcounter{figure}{0}
\renewcommand{\thefigure}{A\arabic{figure}}
\renewcommand{\thetable}{A\arabic{table}}

\input{Filtering_table.tex}

\input{f_table.tex}

\input{IV_results_weights_table.tex}

\blandscape
\input{IV_results_table_alt_measure.tex}
\elandscape
\begin{figure}
\centering 
\includegraphics[width=.9\textwidth,height=.9\textheight,keepaspectratio]{map.jpeg}
\caption{Map of In and Out of Sample Counties}
\end{figure}
